\texttt{vqtl}: An \texttt{R} package for Mean-Variance QTL Mapping

### 2018-05-05 revision/response.tex

\documentclass[11pt]{article} 
\usepackage[letterpaper, headheight=100pt]{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=2.5cm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{xspace}
\newcommand{\eg}{\emph{e.g.}\xspace}
\newcommand{\ie}{\emph{i.e.}\xspace}
%\newcommand{\WV}[2]{\textcolor{red}{#1}\footnote{\textcolor{red}{#2}}}
\newcommand{\WV}[2]{\textcolor{red}{#1\footnote{\textcolor{red}{#2}}}}
\newcommand{\RC}[2]{\textcolor{purple}{#1\footnote{\textcolor{purple}{#2}}}}
\newcommand{\reply}{\ensuremath{\hookrightarrow}\xspace}

\newcommand{\CortyRPaper}{\textcolor{purple}{Corty and Valdar 2018+ [r/vqtl]}\xspace}
\newcommand{\CortyReanalysisPaper}{\textcolor{purple}{Corty \emph{et al} 2018+}\xspace}
\newcommand{\CortyMethodsPaper}{\textcolor{purple}{Corty and Valdar 2018+ [BVH]}\xspace}

@@

\section*{Editor Comments}

\subsection*{Multi-manuscript Issues (identical across responses to reviewers)}

\EditorQuestion{
  Now that you have all three manuscripts potentially going forward for publication in G3, they can be more tightly linked.
}
\Response{
  We have updated the manuscripts such that each one refers to the other two in the introduction and discussion.
}

\EditorQuestion{
   For instance, one of the other papers could provide the benchmark data that is required for [the software] manuscript.
   So the software paper can include those files and refer to the other manuscript for details.
}
\Response{
  We have incorporated this recommendation, using the dataset from Kumar et al. 2013 that was used in one of the case studies in the manuscript titled ``Mean-Variance QTL Mapping Identifies Novel QTL for Circadian Activity and Exploratory Behavior in Mice''.
  The time required to analyze this dataset, according to number of computer cores available and number of permutations desired, is described in the ``benchmarks'' section of the software paper and the relevant files are included in its Zenodo repository.
  (These files are now in two places, since they are also in the Zenodo repository of the reanalysis paper.)
}

@@

\subsection*{Issues Specific to this Manuscript}

\EditorQuestion{
  As suggested by Reviewer 1 it will be very useful to have a benchmark study with times/memory requirements as a function of number of markers and sample size.
  This study will give those interested on using the package an understanding of how this package would scale.
  It will be important to include in the discussion also elements related to scalability: is the package suitable for very large-p/large-n? 
}
\Response{
  We have added this information in the ``Performance Benchmarks'' section.
  See detailed response to Reviewer 1 comment 3.
}

\EditorQuestion{
  There is a very rich literature in the study of genetic control in environmental variance in both plant and animal breeding as well as model organisms.
  This literature should be mentioned. This comment may also apply to some of the companion papers.
  [Citation recommendations provided by the editor.]
}
\Response{
  We have added a section in the introduction that describes how projects in plant and animal breeding seeking to modulate environmental variance are related to the work presented here.
}

@@

% Ibáñez N, Moreno A, Nieto B, Piqueras P, Salgado C, Gutierrez JP. Genetic Parameters Related to Environmental Variability of Weight of Mice; Signs of Correlated Canalised Response. Genet Sel Evol. 2008;40:279-293.

% Ibáñez N, Varona L, Sorensen D, Noguera JL. A study of heterogeneity of environmental variance for slaughter weight in pigs. Animal. 2007;2:19-26.

% Sørensen P, de los Campos G, Morgante F, Mackay TFC, Sorensen D. Genetic Control of Environmental Variation of Two Quantitative Traits of Drosophila melanogaster Revealed by Whole-Genome Sequencing. Genetics. 2015;201(2):487-497. doi:10.1534/genetics.115.180273.

\section*{Reviewer 1 Comments}

\ReviewerQuestion{
  The authors mention ``in almost all traits of interest in human health and disease''.
  The detection of QTL has also been widely used in plant and animal populations.
}
\Response{
  The reviewer is quite correct and we have edited the introduction to reflect the fact that QTL mapping has been applied productively in plant and animal research.
}

\ReviewerQuestion{
  The authors used a set of simulated data with 3 chromosomes and 11 equally spaced markers.
  I would like to see an example with a larger genotype data set.
}
\Response{
  In order to most clearly illustrate the standard workflow associated with package \texttt{vqtl}, the authors have demonstrated its usage on a an unrealistically-small genome.
  The authors are encouraged by the reviewer's enthusiasm to see the package applied to real datasets.
  This manuscript has been submitted as part of a three-paper companionship, where (call this one paper one), paper two contains two real-data reanalyses and paper three contains one real-data reanalysis.
  Thus, the authors contend it is appropriate to maintain this paper having only limited mention of real data, those mentions in the ``Performance Benchmarks'' section.
}

\ReviewerQuestion{
  The CPU time is indicated for the data set of the example.
  I would like to have the CPU time information in a more realistic data set.
  Perhaps authors can use a free data set from the web.
}
\Response{
  Thank you for this recommendation.
  The authors agree that including this information will make the paper more useful to geneticists who are considering using the package -- we have added two sets of benchmarks.
}
\Response{
  First, we demonstrate the performance of the package according to number of computer cores used and number of permutations conducted.
  For this set of benchmarks, we used the intercross from Kumar et. al 2013 which we reanalyzed in \CortyReanalysisPaper, which involved 244 mice, 93 markers, and after using a hidden Markov model to estimate genotype probabilities at locations between markers, 582 loci.
  As recommended by the editor, this example helps tie together the companions, because the results of this analysis are presented in a companion manuscript.
  This addition can be found on page 4.
}
\Response{
  Second, we demonstrate the performance of the package according to the size of the mapping population and the number of genetic markers, holding constant the number of permutations at 1000 and the number of cores at 32.
  For this set of benchmarks, we used crosses simulated with \texttt{R/qtl} so we could control the population size and number of genetic markers.
  This addition can be found on page 4.
}

@@

\ReviewerQuestion{
  In the simulation, the authors should indicate the magnitude of the genetic variance explained by the QTL.
}
\Response{
  Thank you for this recommendation.
  We have added this information to the manuscript and it now appears on page 2.
}

\ReviewerQuestion{
  I think the authors should also indicate if the software is capable in terms of computing power to analyze the data from SNP genotyping devices or even sequencing.
}

\Response{
  Thank you for this recommendation.
  We have addressed this question in the discussion section, on page 4.
}

### 2018-05-05 Package_vqt_Manuscript.tex
responding to reviews

\usepackage{subcaption}

\usepackage{color,xcolor,xspace}
\input{macros}

@@

QTL mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health, and livestock and crop improvement.
Advances in model organism genotyping \citep{Williams1990} and phenotyping \citep{Yang2014a} as well as in statistical methods \citep{Lander1989a,Martinez1992} and software tools (\eg, \citealt{Broman2003,Mulligan2017}) have supported these discoveries.

@@

\section*{Example data: Simulated F2 Intercross and Four Traits}

@@

Each type of test (mQTL, vQTL, and mvQTL) yields two association statistics: the LOD score, and the (nominal) $p$-value. The LOD is a raw measure of association equal to the base 10 logarithm of the likelihood ratio (LR) between the fitted alternative and null models. Higher values indicate greater association when considered across loci for the same type of test; but LOD scores between different types of tests, namely between mvQTL test vs either mQTL or vQTL tests, are not readily comparable. The $p$-value, which is comparable between different types of tests, transforms the LOD score to take account of the number of parameters being fit: it is calculated from the asymptotic distribution of $2\log_e\left(\text{LR}\right)$ under the null model, namely the $\chi^2$ distribution with degrees of freedom equal to the difference in the number of parameters between the alternative and null models. With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.

@@

@@

@@

@@

We have demonstrated typical usage of the \texttt{R} package \texttt{vqtl} for mean-variance QTL mapping in an F2 intercross.
This package is appropriate for crosses and phenotypes where genetic factors or covariates or are known or suspected to influence phenotype variance.
In the case of genetic factors, they can be mapped, as illustrated in \CortyReanalysisPaper.
In the case of covariates, they can be accommodated, which can increase power and control false positive rate, as described in \CortyMethodsPaper.

@@

\begin{figure}[ht!]
    \begin{subfigure}[b]{\linewidth}
        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype1x.pdf}
    \end{subfigure}

@@

\begin{figure}[ht!]
    \begin{subfigure}[b]{\linewidth}
        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype1x.pdf}
    \end{subfigure}

@@

\begin{figure}[htbp]
    \begin{subfigure}[t]{0.5\linewidth}
        \includegraphics[width=\textwidth]{images/mean_var_plot_phen1x.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.5\linewidth}
        \includegraphics[width=\textwidth]{images/mean_var_plot_phen2x.pdf}
    \end{subfigure}

>>>>

\usepackage{subcaption}
\usepackage{multirow}
\usepackage{color,xcolor,xspace}
\input{macros}

@@

QTL mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health, as well as livestock and crop improvement programs.
Advances in model organism genotyping \citep{Williams1990} and phenotyping \citep{Yang2014a} as well as in statistical methods \citep{Lander1989a,Martinez1992} and software tools (\eg, \citealt{Broman2003,Mulligan2017}) have supported these discoveries.

@@

\section*{Example data: Simulated F2 Intercross}

@@

Each type of test (mQTL, vQTL, and mvQTL) yields two association statistics: the LOD score, and the (nominal) $p$-value.The LOD is a raw measure of association equal to the base 10 logarithm of the likelihood ratio (LR) between the fitted alternative and null models. Higher values indicate greater association when considered across loci for the same type of test; but LOD scores between different types of tests, namely between mvQTL test vs either mQTL or vQTL tests, are not readily comparable. The $p$-value, which is comparable between different types of tests, transforms the LOD score to take account of the number of parameters being fit: it is calculated from the asymptotic distribution of $2\log_e\left(\text{LR}\right)$ under the null model, namely the $\chi^2$ distribution with degrees of freedom equal to the difference in the number of parameters between the alternative and null models. With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.

@@

\begin{figure}[t]
    \includegraphics[width = \linewidth]{images/benchmark_kumar.pdf}
    \caption{
        Time taken to run \texttt{scanonevar.perm} on the data from \citet{Kumar2013} which contains 244 individuals and 582 loci, varying the number of permutations desired and the number of computer cores used.
        For a given number of cores, there is a linear relationship between number of permutations conducted and time required.
        The slope the the line indicates time required per permutation and is dependent on the number of cores, ranging from $\approx$ 6.3 seconds per permutation with 4 cores to $\approx$ 1.2 second per permutation with 32 cores.
    }
    \label{fig:benchmark_kumar}
\end{figure}

@@

\begin{figure}[t]
    \includegraphics[width = \linewidth]{images/benchmark_sim_cross.pdf}
    \caption{
        Time taken to run 1000 permutation scans on 32 cores on simulated data using \texttt{scanonevar.perm}, varying the number of individuals in the mapping population and the number of markers in the genome.
        For a given population size, there is a slightly supra-linear relationship between number of markers and time required.
        The average slope of the line indicates the average time required per locus and is dependent on the population size, ranging from $\approx$ 1.4 seconds per locus with a population of size 100 to $\approx$ 3.3 seconds per locus with a population of size 800.
    }
    \label{fig:benchmark_sim_cross}
\end{figure}

@@

\section*{Performance Benchmarks}

By far, the most computationally-intensive step in the mean-variance QTL mapping process is the assessment of genome-wide statistical significance by permutation.
The original genome scan is much faster, because it involves only a single scan, and the bootstrap is much faster because it involves only a single chromosome.

For the first benchmark, we ran \texttt{scanonevar.perm} on the data from \citet{Kumar2013} and \CortyReanalysisPaper, which contains 244 individuals and 582 loci, varying the number of permutations desired and the number of computer cores used.
For a given number of cores, there is a linear relationship between number of permutations conducted and time required \autoref{fig:benchmark_kumar}.
The slope the the line indicates time required per permutation and is dependent on the number of cores, ranging from $\approx$ 6.3 seconds per permutation with 4 cores to $\approx$ 1.2 second per permutation with 32 cores.

For the second benchmark, we ran \texttt{scanonevar.perm} on simulated data, always conducting 1000 permutations and using 32 cores, but varying the number of individuals in the mapping population and the number of markers in the genome.
For a given population size, there is a slightly supra-linear relationship between number of markers and time required \autoref{fig:benchmark_sim_cross}, which reflects a linear increase in the time taken to conduct the permuted genome scans, plus an increase in the time taken for ``bookkeeping'' tasks like organizing and reshaping genetic data.
The slope of the line indicates the time required per locus and is dependent on the population size, ranging from $\approx$ 1.4 seconds per locus with a population of size 100 to $\approx$ 3.3 seconds per locus with a population of size 800.

% Performance benchmarks of package \texttt{vqtl} are presented in \autoref{fig:benchmark_kumar} and \autoref{fig:benchmark_sim_cross}.

% \begin{table}
%     \centering
%     \begin{tabular}{r|rrrr}
%         \multirow{2}{1in}{\raggedleft \textbf{number of permutations}}& \multicolumn{4}{c}{\textbf{number of cores}}\\
%         \cline{2-5}
%         & 4 & 8 & 16 & 32\\
%         \hline
%         100  & 11 & 6 & 4 & 4\\
%         1000 & -  & 53 & 34 & 20\\
%         \hline
%     \end{tabular}
%     \caption{
%         Time in minutes required to conduct permuations for all three tests on the dataset from \citet{Kumar2013}, which has 244 individuals and 582 loci, according to number of permutations desired and number of computer cores available.
%         We recommend 1000 permutations for publication-quality precision.
%         We include the time taken to conduct 100 and 500 permutations to illustrate the linear relation between number of permutations and time required.
%         This table also illustrates the linear relationship between number of computer cores used and time taken to complete permutations.
%     }
% \end{table}
%
% \begin{table}
%     \centering
%     \begin{tabular}{r|rrrr}
%         \multirow{2}{1in}{\raggedleft \textbf{number of markers}}& \multicolumn{4}{c}{\textbf{number of individuals}}\\
%         \cline{2-5}
%         & 100 & 200 & 400 & 800\\
%         \hline
%         100 & 2 & 3 & 4 & 4\\
%         200 & 5 & 5 & 7 & 9\\
%         400 & 9 & 11 & 13 & 19\\
%         800 & 19 & 23 & 31 & 46\\
%         \hline
%     \end{tabular}
%     \caption{Time in minutes required to conduct 1000 permuations on 32 cores for all three tests on a simulated F2 intercross, according to number of individuals in the mapping population and number of markers in the genome.}
% \end{table}

@@

We have demonstrated typical usage of the \texttt{R} package \texttt{vqtl} for mean-variance QTL mapping in an F2 intercross.
This package is appropriate for crosses and phenotypes where genetic factors or covariates or are known or suspected to influence phenotype variance.
In the case of genetic factors, they can be mapped, as illustrated in \CortyReanalysisPaper.
In the case of covariates, they can be accommodated, which can increase power and improve false positive rate control, as described in \CortyMethodsPaper.

@@

\begin{figure}[ht]
    \begin{subfigure}[b]{\linewidth}
        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype1x.pdf}
    \end{subfigure}

@@

\begin{figure}[ht]
    \begin{subfigure}[b]{\linewidth}
        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype1x.pdf}
    \end{subfigure}

@@

\begin{figure}[ht]
    \begin{subfigure}[t]{0.5\linewidth}
        \includegraphics[width=\textwidth]{images/mean_var_plot_phen1x.pdf}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.5\linewidth}
        \includegraphics[width=\textwidth]{images/mean_var_plot_phen2x.pdf}
    \end{subfigure}
